---
title: "Lecture 12"
subtitle: "Logistic regression"
author: "Jung-Jin Lee"
date: "Apr 20, 2021"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Drexel.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
 
```{r, message = F, echo = F, warning = F}
library(tidyverse) 
library(gridExtra)
library(knitr) 
library(latex2exp)
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 4,
  fig.height = 4
  )
filter <- dplyr::filter
options(width = 70) # controls console (and slide) output width
```

## Odds

As a measure of likelihood of an event, the **odds** of an event are the ratio of the probability that the event occurs to the probability that the event cannot occur. In other words, for an event $A$, the odds of $A$, denoted $\text{odds}(A)$, is defined to be 
$$\text{odds}(A) = \frac{P(A)}{1-P(A)}.$$

**Example**: when rolling a fair die, the odds of getting 1 are $\frac{1/6}{5/6} = \frac{1}{5}$ (or 1:5). The odds of getting an even number are $\frac{1/2}{1/2} = 1$ (or 1:1).

--

It is notable that $P(A)$ can be recovered from $\text{odds}(A)$:
$$P(A)=\frac{\text{odds}(A)}{1+\text{odds}(A)}.$$
---

## Titanic dataset

.pull-left-code[
- On April 15, 1912, the largest passenger liner ever made collided with an iceberg during her maiden voyage. 

- When the Titanic sank it killed 1502 out of 2224 passengers and crew. 
  - One of the reasons that the shipwreck resulted in such loss of life was that there were not enough lifeboats for the passengers and crew.
  
  - Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others.
]
.pull-right-plot[
<p align = "center">
  <img src = Titanic.png width = "240">
</p>  
]

---

## Titanic dataset

Download a comma separated file [titanic.csv](titanic.csv). It consists of 887 observations and 8 variables. You can read in the dataset using `read_csv()`, a function that comes with `tidyverse`. 

```{r}
titanic_ori <- read_csv(file = "titanic.csv") #<<
head(titanic_ori)
```

---

## Titanic dataset

```{r}
names(titanic_ori)
```

- `Survived`: did the passenger survive? (1 = survived)

- `Pclass`: passenger-class

- `Siblings/Spouses Aboard`: number of siblings/spouses aboard

- `Parents/Children Aboard`: number of parents/children aboard

- `Fare`: fare in UK pounds

--

```{r}
titanic <- titanic_ori %>%
  mutate(Pclass = factor(Pclass, levels = c(1, 2, 3))) %>%
  mutate(Sex = factor(Sex, levels = c("female", "male")))
```

---

## Titanic dataset

`Age` distribution by `Sex` and `Pclass`:

.pull-left-40[
```{r, eval = F}
titanic %>% 
  ggplot(aes(Sex, Age)) + 
  geom_boxplot() + 
  facet_wrap(~Pclass) 
```
]

.pull-right-60[
```{r, echo = F}
titanic %>% 
  ggplot(aes(Sex, Age)) + 
  geom_boxplot() + 
  facet_wrap(~Pclass)
```
]

---

## Logistic regression

How does `Age` affect the `Survived`? 

--

.pull-left-55[
```{r, eval = F}
fit.lm <- lm(Survived ~ Age, 
             data = titanic)
beta0 <- fit.lm$coefficients[1]
beta1 <- fit.lm$coefficients[2]
ggplot(titanic, aes(Age, Survived)) +
  geom_point() +
  geom_abline(intercept = beta0, 
              slope = beta1, 
              color = "red")
```
]

.pull-right-45[
```{r, echo = F}
fit.lm <- lm(Survived ~ Age, 
             data = titanic)
beta0 <- fit.lm$coefficients[1]
beta1 <- fit.lm$coefficients[2]
ggplot(titanic, aes(Age, Survived)) +
  geom_point() +
  geom_abline(intercept = beta0, 
              slope = beta1, 
              color = "red") 
```
]

--

A simple linear regression does not work properly -- for example, what can be said about the survival of a 40-year-old passenger?

---

## Logistic regression

In logistic regression, we model 
$$\log(\text{Odds of survival}) = \beta_0 + \beta_1\texttt{Age},$$
--
that is, 
$$\text{Odds of survival} = e^{\beta_0 + \beta_1\texttt{Age}},$$
or equivalently,
$$\frac{P(\text{Survival})}{1-P(\text{Survival})} = e^{\beta_0 + \beta_1\texttt{Age}}.$$
--
This can be rewritten as 
$$P(\text{Survival}) = \frac{e^{\beta_0 + \beta_1\texttt{Age}}}{1+e^{\beta_0 + \beta_1\texttt{Age}}}.$$
--
The coefficients $\beta_0$ and $\beta_1$ are estimated by the function `glm()`. 

---

## Logistic regression

```{r}
fit.glm <- glm(Survived ~ Age, family = binomial(link = "logit"), #<<
               data = titanic) #<<
fit.glm$coefficients
```

```{r, echo = F}
beta0 <- fit.glm$coefficients[1]
beta1 <- fit.glm$coefficients[2]
```

From this output, we conclude that

$$\log(\text{Odds of survival}) = -0.209188757-0.008774355 \times \texttt{Age}$$

--

For a 40-year-old passenger, the odds of survival is
$$e^{-0.209188757-0.008774355 \times 40} = 0.571116.$$
In other words, the probability of survival is
$$\frac{e^{-0.209188757-0.008774355 \times 40}}{1+e^{-0.209188757-0.008774355 \times 40}}=36.35\%.$$

---

## Interpretation of coefficients

Interpretation of $\beta_1$ (-0.008774355 in the previous example):
--

- Odds of survival at age $x$:
$$\text{Odds}(x) = e^{\beta_0 +\beta_1x}.$$
--

- Odds of survival at age $x+1$ (a unit increase in age):
$$\text{Odds}(x+1) = e^{\beta_0 +\beta_1(x+1)}.$$
--

- In other words, 
$$\text{Odds}(x+1) = e^{\beta_0 +\beta_1(x+1)} = e^{\beta_0 +\beta_1x+\beta_1}=e^{\beta_1}\times\text{Odds}(x),\quad\text{or}$$
$$\frac{\text{Odds}(x+1)}{\text{Odds}(x)}=e^{\beta_1}.$$
--


In general, $e^{\beta_1}$ represents the ratio of two odds when there is a unit change in the predictor variable. 

---

## Interpretation of coefficients

$$\frac{\text{Odds}(x+1)}{\text{Odds}(x)}=e^{\beta_1}.$$

--

- $\beta_1 > 0 \Leftrightarrow e^{\beta_1} > 1 \Leftrightarrow \text{Odds}(x+1) > \text{Odds}(x) \Leftrightarrow$ increase in the predictor leads to increase in the odds.

--

- $\beta_1 < 0 \Leftrightarrow e^{\beta_1} < 1 \Leftrightarrow \text{Odds}(x+1) < \text{Odds}(x) \Leftrightarrow$ increase in the predictor leads to decrease in the odds.

---

## Logistic regression with a categorical predictor

How does `Sex` affect the `Survived`? 

```{r}
fit.glm2 <- glm(Survived ~ Sex, family = binomial(link = "logit"),
                data = titanic)
fit.glm2$coefficients
```

--

Interpretation:

- The odds of survival of female = $e^{\beta_0} = e^{1.056589} = 2.877$. This implies that the probability of survival of a female passenger equals $\frac{e^{1.056589}}{1+e^{1.056589}}=74.2\%$. 

--

- On the other hand, the odds of survival of male = $e^{\beta_0 + \beta_1} = e^{1.056589-2.505126} = 0.235$. This implies that the probability of survival of a female passenger equals $\frac{e^{1.056589-2.505126}}{1+e^{1.056589-2.505126}}=19.0\%$.

---

## Logistic regression with a categorical predictor

```{r, out.width = "50%"}
titanic %>%
  mutate(Survival = ifelse(Survived == 1, "yes", "no")) %>%
  ggplot(aes(Sex, fill = Survival)) +
  geom_bar()
```

---

## Logistic regression with a categorical predictor

```{r}
table(titanic$Survived, titanic$Sex)
female_survival_prob <- 233/(81+233)
male_survival_prob <- 109/(464+109)
female_survival_prob; male_survival_prob
```

---

## Logistic regression with multiple predictors

How do `Age` and `Pclass` affect the `Survived`? 

```{r}
fit.glm3 <- glm(Survived ~ Age + Pclass, 
                family = binomial(link = "logit"),
                data = titanic)
fit.glm3$coefficients
```

--

Interpretation:

- When `Pclass` remains unchanged, the odds of survival of a patient with age $x+1$ is $e^{-0.03946933}=0.9612994$ times the odds of survival of a patient with age $x$ (The older, the less likely to survive).

--

- When `Age` remains unchanged, the odds of survival of a patient in passenger-class $2$ is $e^{-1.03811692}=0.3541209$ times the odds of survival of a patient in passenger-class $1$.

--

- When `Age` remains unchanged, the odds of survival of a patient in passenger-class $3$ is $e^{-2.28986274}=0.1012804$ times the odds of survival of a patient in passenger-class $1$.

---

## Logistic regression with multiple predictors

```{r, out.width = "50%"}
titanic %>%
  mutate(Survival = ifelse(Survived == 1, "yes", "no")) %>%
  ggplot(aes(Pclass, fill = Survival)) +
  geom_bar()
```

---

## Exercise

Recall the horseshoe crabs dataset. We want to model the existence of satellites of a female crab using the weight and spine condition of the  female crab.

```{r}
crabs <- read_tsv("crabs.tsv") %>%
  mutate(has.satell = ifelse(satell > 0, 1, 0)) %>%
  mutate(spine = factor(spine, levels = c("1", "2", "3"))) %>%
  mutate(weight = weight/1000) # weight in grams to kilograms
```
--

- What is the probability that a female crab with weight 4.1kg and spine condition 3 has at least one satellite?
- What is the probability that a female crab with weight 2.8kg and spine condition 1 has at least one satellite?

---

## Exercise

```{r}
fit.crab <- glm(has.satell ~ weight + spine,
                family = binomial(link = "logit"),
                data = crabs)
fit.crab$coefficients
```
--

For female crabs with spine condition 3, the odds of having a satellite equals 
$$e^{-3.65268080 + 1.78814346\times\texttt{weight}+0.07082751}.$$
--

In particular if the weight is $4.1$, then the odds become
$$e^{-3.65268080 + 1.78814346\times4.1+0.07082751}=42.50.$$
--

This implies that the probability of having a satellite is 
$$\frac{e^{-3.65268080 + 1.78814346\times4.1+0.07082751}}{1+e^{-3.65268080 + 1.78814346\times4.1+0.07082751}}=97.7\%.$$

---

## Exercise

For female crabs with spine condition 1, the odds of having a satellite equals 
$$e^{-3.65268080 + 1.78814346\times\texttt{weight}}.$$
--

In particular if the weight is $2.8$, then the odds become
$$e^{-3.65268080 + 1.78814346\times2.8}=3.87.$$
--

This implies that the probability of having a satellite is 
$$\frac{e^{-3.65268080 + 1.78814346\times2.8}}{1+e^{-3.65268080 + 1.78814346\times2.8}}=79.5\%.$$

---

## Exercise

```{r, out.width = "50%"}
crabs %>%
  mutate(has.satell = ifelse(has.satell == 1, "yes", "no")) %>%
  ggplot(aes(spine, fill = has.satell)) +
  geom_bar()
```

---

## Exercise

```{r, out.width = "50%"}
crabs %>%
  mutate(has.satell = ifelse(has.satell == 1, "yes", "no")) %>%
  ggplot(aes(has.satell, weight)) +
  geom_boxplot()
```

---

## Final exam

- Tuesday, April 27. 5:30pm ~ 10:00pm

- PDF preferred

- Cumulative (more than 70% from material after Exam 2)

--

```{r}
num1 <- c(20, 8, 1, 14, 11)
num2 <- c(25, 15, 21)
char1 <- LETTERS[num1]
char2 <- LETTERS[num2]
IWannaSay <- paste(paste(char1, collapse = ""),
                   paste(char2, collapse = ""))
print(IWannaSay)
```


```

