---
title: "Homework 11"
subtitle: "due Apr 20, 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center")
library(tidyverse)
```

For all questions below, consider the `crabs` data set that we used in class:

```{r, echo = T, eval = T}
crabs <- read.table(file = "crabs.tsv", header = T, sep = "\t")
dim(crabs); names(crabs)
```

Consider the following three models for `satell`:

$$\textbf{Model 1: } \texttt{satell} = \beta_0 + \beta_1 \texttt{width},$$
$$\textbf{Model 2: } \texttt{satell} = \beta_0 + \beta_1 \texttt{weight},$$
$$\textbf{Model 3: } \texttt{satell} = \beta_0 + \beta_1 \texttt{width} + \beta_2 \texttt{weight}.$$

```{r, echo = F, eval = F}
crabs <- read_tsv("crabs.tsv")
fit1 <- lm(satell ~ width, data = crabs)
fit2 <- lm(satell ~ weight, data = crabs)
fit3 <- lm(satell ~ width + weight, data = crabs)

summary(fit1)$r.squared
summary(fit2)$r.squared
summary(fit3)$r.squared
```

**1.** Compute $R^2$ for each model. 

```{r, echo = F, eval = F}
AIC(fit1, fit2, fit3)
anova(fit1, fit3)
anova(fit2, fit3)
```

\bigskip

**2.** Which model is the best? Justify your answer. 

```{r}
crabs <- crabs %>%
  mutate(color = factor(color, levels = c(5,4,3,2))) %>%
  mutate(spine = factor(spine))
```

\bigskip

**3.** Create the following box plots. Note the order of the labels in the $x$-axis. _Hint_: change `color` to a factor variable with levels 5, 4, 3, 2, in this order. 

```{r, fig.height=3}
crabs %>%
  ggplot(aes(color, weight)) +
  geom_boxplot() +
  theme(aspect.ratio = 1)
```

\newpage

**4.** Carry out an appropriate regression between `weight` (response) and `color` (explanatory). How does the change in color from `5` to `3` affect `weight`?

```{r}
fit <- lm(weight ~ color, data = crabs)
```

**5.** How does the change in color from `2` to `4` affect `weight`? _Hint_: you might want to change `color` to a factor variable with levels 2, 3, 4, 5, in this order.

```{r}
fit <- lm(weight ~ color, data = crabs)
```


