---
title: "Homework 9"
subtitle: "due Apr 6, 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center")
library(tidyverse)
```

(For questions **1, 2, 3**) Consider the following paired data sets of length $20$:

```{r, echo = T}
x <- c(6.82, 1.44, 9.39,  8.51, 10.38,  4.59, 14.96,  9.68, 13.54,  6.42, 11.03,  
       3.53, 16.91,  9.52,  8.16,  8.97,  8.32,  3.58, 13.57,  9.99)
y <- c(36.69, 6.39, 49.59, 45.65, 52.18, 27.66, 79.35, 54.10, 71.01, 34.60, 61.17, 
       22.79, 91.20, 50.57, 44.11, 53.51, 45.96, 22.20, 73.01, 55.70)
```

**1.** Create a scatter plot to visualize the data (_Hint_: you may want to start with making a data frame and then use `geom_point()`, `x` as $x$-axis and `y` as $y$-axis). 

```{r, echo = F, eval = F}
df <- data.frame(x, y)
g <- ggplot(df, aes(x, y)) + geom_point()
print(g)
```

**2.** Do you think there is a strong linear association between `x` and `y`? Compute the sample correlation coefficient between `x` and `y` to justify your claim. 

```{r, echo = F, eval = F}
cor(x, y)
```

**3.** Assume that `y` is an outcome in a certain experiment, and `x` is a predictor. Find the best fitting
line describing the association between `x` and `y` by specifying its $y$-intercept ($\beta_0$) and slope ($\beta_1$). Overlay the best fitting line to the plot you obtained in **1.** above.

```{r, echo = F, eval = F}
fit <- lm(y ~ x)
fit$coefficients
```

\bigskip

(For questions **4, 5**) Consider the `mtcars` data set we used in class:

```{r, echo = T, eval = T}
head(mtcars, n = 3)
```

**4.** Obtain the following scatter plot with the best fitting line:

```{r, echo = F, eval = T, fig.height = 2.5}
fit <- lm(hp ~ wt, data = mtcars)
beta0 <- fit$coef[1]; beta1 <- fit$coef[2]
ggplot(mtcars, aes(wt, hp)) + 
  geom_point() + 
  geom_abline(intercept = beta0, slope = beta1, color = "red") +
  theme(aspect.ratio = 1)
```

**5.** Compute $r$, the correlation coefficient. 



