---
title: "Quadratic equation solver"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Make a function `quad_sol()` that solves a given quadratic equation. For example, to solve
$$2x^2+5x-3=0,$$
the function takes three coefficients $2$, $5$, and $-3$ and `quad_sol(2, 5, -3)` should return the roots $0.5$ and $-3$. Make sure that `quad_sol(a, b, c)` can take care of the case when $a = 0$ (where the given equation is in fact linear). Also `quad_sol` should be able to handle the cases of imaginary roots and double root by returning a sentence like `No real roots` and `There is a double root 2`.  Some examples follow:

```{r, echo = F}
lin_sol <- function(a, b) {
  if (a == 0) {
    if (b == 0) {
      print("Infinitely many solutions")
    } else {
      print("No solutions")
    }
  } else {
    y <- -b/a
    re <- paste0("It has a single root ", y)
    print(re)
  }
}
```

```{r, echo = F}
quad_sol <- function(a, b, c)  {
  if (a == 0) {
    lin_sol(b, c)
  }
  else {
    D <- b^2-4*a*c
    if (D > 0) {
      y1 <- (-1*b + sqrt(D)) / (2*a)
      y2 <- (-1*b - sqrt(D)) / (2*a)
      re <- paste0("There are two real roots ", y1, " and ", y2)
      print(re)
    } else if (D ==0) {
      y <- -1*b / (2*a)
      re <- paste0("There is a double root ", y)
      print(re)
    } else {
        print("No real roots")
    }
  }
}
```

```{r}
quad_sol(1, 2, 1)
quad_sol(2, 5, -3)
quad_sol(2, 5, 13)
quad_sol(0, 3, 11)
quad_sol(0, 3, 0)
quad_sol(0, 0, 1)
```

