---
title: "Philadelphia temperature"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Download a [file](2249105.csv) which contains Philadelphia International Airport daily temperature data from Jan 1, 2020 through Aug 13, 2020.

```{r, echo = F}
d <- read.csv("2249105.csv") %>%
  select(NAME, DATE, TMAX, TMIN) %>%
  filter(NAME == "PHILADELPHIA INTERNATIONAL AIRPORT, PA US") %>%
  mutate(DATE = as.Date(DATE))
```

- Read in the data file and select only `NAME`, `DATE`, `TMAX` (high temperature), and `TMIN` (low temperature). Only leave Philadelphia data. Make sure that `DATE` is a date variable. 

- Create the following graph.

```{r, echo = F}
d %>% 
  ggplot(aes(x = DATE)) +
  geom_line(aes(y = TMAX), color = "red") +
  geom_line(aes(y = TMIN), color = "blue") +
  theme_bw()
```

- Modify the graph above to get the following graph with a legend. 

```{r, echo = F}
d %>% 
  pivot_longer(cols = c("TMAX", "TMIN"), names_to= "low_high", values_to = "Temperature") %>%
  ggplot(aes(x = DATE, y = Temperature, color = low_high)) +
  geom_line() +
  theme_bw()
```

