---
title: "Lecture 2"
subtitle: "Exploring data through visualization"
author: "Jung-Jin Lee"
date: "Jan 14, 2020"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "Drexel.css"] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---
 
```{r, message = F, echo = F, warning = F}
library(tidyverse) 
library(gridExtra)
library(knitr) 
opts_chunk$set(
  tidy = FALSE,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 4,
  fig.height = 4
  )
options(width = 65) # controls console (and slide) output width
```

## Type of data

- Categorical Data - consist of qualitative values
  - Nominal values: gender, coat colors
  - Ordinal values: body conditions, age groups
  
- Numerical Data - consist of quantitative values
  - Discrete values: litter size, parity
  - Continuous values: height, weight, temperature
  
---

## Exploring data

```{r}
library(tidyverse) # need to load the tidyverse library #<<
dmd <- diamonds # diamonds comes with ggplot2
dim(dmd)
head(dmd)
```

---

## Exploring data, continued

How many distinct values are there in the variable `cut`?

```{r}
DiaCut <- dmd$cut
unique(DiaCut) #<<
```

--

What is the frequency of each value?

```{r}
table(DiaCut) #<<
```

---

## Bar graph

One way to visualize categorical data is to use a bar graph

```{r, out.width = "50%"}
g <- ggplot(data = dmd, aes(x = cut)) + #<<
  geom_bar() #<<
print(g)
```

---

## Exercise

- How many distinct values are there in the variable `color`?
- What is the frequency of each value in `color`?
- Draw a bar graph showing the frequency

--

```{r}
unique(dmd$color)
```

--

```{r}
table(dmd$color)
```

---

## Exercise, continued

```{r, out.width = "50%"}
ggplot(dmd, aes(color)) + ## can omit 'data = ' and 'x = '
  geom_bar()
```

---

## Two-way table

Frequency of color for each cut

```{r}
table(dmd$cut, dmd$color)
```

---

## Graphs using facets

```{r, out.width = "50%"}
ggplot(dmd, aes(color)) + ## can omit 'data = ' and 'x = '
  geom_bar() +
  facet_wrap(~cut) #<<
```

---

## Colored bar graphs

```{r, out.width = "75%", fig.asp = 0.7}
ggplot(dmd, aes(color, fill = clarity)) + #<<
  geom_bar() +
  facet_wrap(~cut) 
```

---

## Exercise

Produce the following plots:

```{r, echo = F, out.width = "75%", fig.asp = 0.85}
ggplot(dmd, aes(cut, fill = clarity)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5)) +
  facet_wrap(~color) 
```

---

## Description of numerical data

What is the distribution of `carat` like?

```{r, out.width = "50%"}
ggplot(dmd, aes(carat)) +
  geom_histogram() #<<
```

---

## Location and spread of numerical data

- Measure of location: sample mean and median
- Measure of spread: sample variance and standard deviation

---

## Example: find the sample mean

The sample mean is given by sum of samples divided by the number of samples:
$$\bar{x} = \frac{x_1+x_2+\cdots+x_n}{n}$$

```{r}
ht <- read.table(file = "heights.txt", header = TRUE, sep = " ")
mom <- ht$Mheight
s <- sum(mom); n <- length(mom)
mom_height_mean <- s/n
print(mom_height_mean)
```

R has a built in function to compute the mean:

```{r}
mean(mom) # mean() gives the mean of a vector #<<
```

---

## Sample median

The median is the value separating the higher half from the lower half of numerical data. R has a built in function to compute the median. 

```{r}
median(mom) #<<
```

In general, one can compute quantiles (cut points dividing range of numerical data).

```{r}
quantile(mom) #<<
```

The median is the same as the 50% quantile. 

---

## Example

**Exercise**: Find the sample mean and median of `carat`.

--

```{r}
carat <- dmd$carat
mean(carat); median(carat)
```

--

**Exercise**: Find the 90% quantile `carat`.

--

```{r}
quantile(carat) 
```

--

```{r}
quantile(carat, probs = 0.9) #<<
```

---

## Sample variance

The sample variance measures spread of given data:
$$s^2 =\frac{(x_1-\bar{x})^2+(x_2-\bar{x})^2+\cdots+(x_n-\bar{x})^2}{n-1}.$$
Note that the denominator is $n-1$, not $n$.

--

One could also compute the sample variance using the definition:

```{r}
v1 <- mom-mean(mom); v2 <- v1^2
sum(v2)/(length(mom)-1)
```

--

R has a built in function to compute the sample variance:

```{r}
var(mom) # var() gives the sample variance of a vector #<<
```

---

## Sample standard deviation

The sample standard deviation also measures spread of given data -- it is simply the square root of the sample variance:
$$s = \sqrt{s^2} = \sqrt{\frac{(x_1-\bar{x})^2+(x_2-\bar{x})^2+\cdots+(x_n-\bar{x})^2}{n-1}}.$$

R has a built in function to compute the sample standard deviation:

```{r}
sd(mom) # sd() gives the sample standard deviation of a vector #<<
```

---

## Exercise

- Produce the histogram of daughter's heights from the dataset `heights.txt`.

- Get mean/median/sample variance of daughter's heights. 

---

## Exercise, continued

```{r, out.width = "40%"}
ggplot(ht, aes(Dheight)) + geom_histogram()
```

```{r}
D_height <- ht$Dheight
c(mean(D_height), median(D_height), var(D_height))
```

---

## Box plot

**Example**: using `diamonds` data set, describe the distribution of `carat` for each `cut`

```{r, out.width = "45%"}
g <- ggplot(dmd, aes(x = cut, y = carat)) +
  geom_boxplot() #<<
print(g)
```

---

## Box plot, continued

- The thick line indicates the median.
- The lower and upper hinges correspond to the 25% and 75% quantiles. 
- The upper whisker extends from the hinge to the largest value no further than 1.5 $\times$ IQR from the hinge, where IQR is the inter-quartile range (distance between the 25% and 75% quantiles). Similarly for the lower whisker.

---

## Violin plot

**Example**: using `diamonds` data set, describe the distribution of `carat` for each `cut`

```{r, out.width = "45%"}
g <- ggplot(dmd, aes(x = cut, y = carat)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) #<<
print(g)
```

---

## Scatter plots

**Example**: how are `carat` and `price` associated in `diamonds` data set?

```{r, out.width = "50%"}
g <- ggplot(dmd, aes(x = carat, y = price)) +
  geom_point()
print(g)
```

---

## Scatter plots, continued

```{r, fig.width = 6, fig.height = 4.5, out.width = "70%"}
g <- ggplot(dmd, aes(x = carat, y = price, color = cut)) +
  geom_point() 
print(g)
```

---

## Improving plots

```{r, fig.width = 6, fig.height = 4.5, out.width = "60%"}
g <- ggplot(dmd, aes(x = carat, y = price, color = cut)) +
  geom_point() +
  theme_classic() + #<<
  labs(x = "Carats", y = "Price of diamond", color = "Cut values") + #<<
  ggtitle("Diamonds data") #<<
print(g)
```

---

## Exporting plots

`ggsave()` exports plots generated by `ggplot`:

```{r, eval = F}
ggsave(filename = "diamonds.pdf", g) #<<
```

--

Other file formats are available:

```{r, eval = F}
ggsave(filename = "diamonds.png", g) 
ggsave(filename = "diamonds.jpeg", g) 
```
