---
title: "Homework 4"
subtitle: "due Feb 9, 2021"
output:
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center")
library(tidyverse)
```

\bigskip

(Problems **1**, **2**) Recall the `diamonds` data set we used in class.

**1.** Note that the `diamonds` data set has eight distinct `clarity` values. For each of these values, compute the median diamond `price`.

**2.** Verify your answer in **1.** above by drawing the plot below:

```{r, echo = F, fig.width = 3.3, fig.height = 3}
diamonds %>%
  ggplot(aes(clarity, price)) +
  geom_boxplot()
```

**3.** Recall the `starwars` data set we used in class. Which human female character has the shortest height with brown hair color?

**4.** Assume that $X$ follows a normal distribution with mean 5 and standard deviation 2, say $N(3,2^2)$. Compute the probability that $3\leq X \leq 5$.

**5.** Determine $c$ so that the probability $0\leq Z \leq c$ is $0.45$, where $Z$ follows the standard normal distribution $N(0,1^2)$.